---
title: "Life history theory"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstudioapi)
current_path <- getActiveDocumentContext()$path 
#setwd(dirname(current_path)) # set working directory to location of this file
```


## Learning objectives

1. Know the terms and reasons for analyzing structured populations
2. Convert life history information into a matrix for a stage-structured population
3. Get population growth rate from a matrix

*Review of last lecture: environmental variability and how to deal with that*

*modeled population growth with seed bank*

*assumed all individuals are same - in reality, different stages & ages within population*

*Silvertown talked about tradeoffs between growth, survival and fecundity, we'll talk about this more next week*

*built his analyses with structured population models, so introducing that first this week*

## Structured populations

* structured populations: accounting for stage or age of individuals *stage examples: seeds, vegetative, reproductive*

*question: any ideas why it might be useful to incorporate this information for a population?*

* differences in mortality more realistic

*example: differences in mortality, seedlings die in frost, reproductive adults don't*

* individual stages - consequences for population dynamics 

*population growth in a population of only seedlings will look different than in population of older adults*

### Terminology

* demography - study of structured populations
* vital/demographic rates - germination, survival, fecundity *rates that describe each stage*

*vital & demographic used interchangably*

* age and stage *related but useful in different circumstances*
* stable age/stage distribution - proportion of individuals in each age/stage 

## Matrix population models

*let's say we have a plant with 3 stages: seed, seedling, and adult*

*we're interested in getting germination rates, survival rates, fecundity*

*question: any ideas how might we actually collect that data in a real population with a bunch of individuals in seed, seedling, adult stage?*

* data collection: counts in each stage, tracking individuals
* rates of transition between each stage

*no seed bank, all seeds are coming from adults in previous year*

| Year | seeds | seedlings | adults |
| :----: | :----: | :----: | :----: |
| 1 | 10 | 10 | 10 |
| 2 | 40 | 1 + 2 | 8 + 3 | 

*in year one, we tag seeds, seedlings, and adults, and measure fecundity of adults*

*in year two, we have seeds from last year's adults, 1 previously tagged seedling, 2 new seedlings, 8 previously tagged adults, 3 new adults*

*quick aside: has this population reached its stable stage distribution? in other words, are the percentages of seeds, seedlings, and adults consistent from one year to the next?*

| Year | seeds | seedlings | adults |
| :----: | :----: | :----: | :----: |
| 1 | 33% | 33% | 33% |
| 2 | 74% | 6% | 20% | 

* initial population =/= long term dynamics
* transient period
* number of individuals not constant, proportion in each stage constant

*returning to matrix population model*

*this information from 1 year transition is all we need to create a matrix population model, assuming stable stage distribution*

* each column = stage
* column 1 = rates from seeds to seeds/seedlings/adults
* column 2 = seedlings to seeds/seedlings/adults
* column 3 = adults
* adults to seeds is fecundity

```{r}
stage <- c(0,0.2,0,0,0.1,0.3,4,0,0.8)
matrix(stage,nrow=3,ncol=3)
```

*label columns & rows*

*can think of 0s as percentage of one stage going to others*

*for example, [3,1] is 0 because no seeds are going straight to adults*

*[1,2] is 0 because no seedlings are going back to seeds*

### Calculating $\lambda$

*use linear algebra, not going to get into that, just some definitions*

* eigenvalue = a scalar used in matrix multiplication
* several eigenvalues for a given matrix *i.e. several scalars can work when doing this transformation/matrix multiplication*
* largest, positive eigenvalue of matrix = $\lambda$ = population growth rate

*really easy to do in R, will do in lab*

## (maybe skip) Age-structured populations: Leslie matrix

*let's put this in the leslie matrix format*

* key difference: can't stay in an age *like you can in a stage*
* fecundity in first row
* each age is a column
* survival to next stage on the diagonal

```{r}
leslie <- c(0,0.2,0,0,0,0.25,100,0,0)
matrix(leslie,nrow=3,ncol=3)
```

