---
title: "Life history theory"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rstudioapi)
current_path <- getActiveDocumentContext()$path 
#setwd(dirname(current_path)) # set working directory to location of this file
```

## Learning objectives

1. Know the terms and methods for analyzing structured populations
2. Convert life history information into a Leslie matrix
3. Get population growth rate from a Leslie matrix

*Review of last lecture: environmental variability and how to deal with that*

*modeled population growth with seed bank*

*assumed all individuals are same - in reality, different stages & ages within population*

## Population structure

### Terminology

* demography - structured populations
* vital rates - describe each stage *germination, survival to reproduction, fecundity*
* age and stage *related but useful in different circumstances*

*stage examples: seeds, vegetative, reproductive*

### Why care about population structure?

* more realistic *that births and deaths are age/stage structured*

*example: seedlings die in frost, reproductive adults don't*

* different stages, different pressures *seeds-soil conditions, adults-competition*
* compare individuals within stage

## Modeling life history stages

* life tables - rates at each stage
* abundance, survival, and fecundity

*let's say we have a plant with 3 stages: seed, seedling (not reproductive), and adult*

*seeds are in a seed bank*

| Stage | Description | Number Alive | Survival to Next Stage | Mortality | Fecundity |
| ----- | ----------- | ----------------: | --------------------------: | -------------: | -------------: |
| 1     | Seed        |              1000 |                        0.20 |           0.80 |              0 |
| 2     | Seedling    |               200 |                        0.25 |           0.75 |              0 |
| 3     | Adult       |                50 |                           — |              — |            100 |

* convert to matrix 

## Leslie matrix

* quantify and analyze structured populations
* inputs: survival and fecundity rates from each age
* calculate population growth rate and project abundance in each age 

*let's take the stage-structured life table from last lecture and make it a biennial*

*now ages instead of stages*

| Age | Description | Number Alive | Survival to Next Stage | Fecundity |
| ----- | ----------- | ----------------: | --------------------------: | -------------: |
| 0     | Seed        |              1000 |                        0.20 |              0 |
| 1     | Seedling    |               200 |                        0.25 |              0 |
| 2     | Adult       |                50 |                           — |            100 |

*let's put this in the leslie matrix format*

* fecundity in first row
* each age is a column
* survival of each age on the diagonal

```{r}
leslie <- c(0,0.2,0,0,0,0.25,100,0,0)
matrix(leslie,nrow=3,ncol=3)
```

*can think of 0s as percentage of one stage going to others*

*for example, [3,1] is 0 because no seeds are going straight to adults*

*[2,2] is 0 because no seedlings are going back to seeds*

### Calculating $\lambda$

*use linear algebra, not going to get into that, just some definitions*

* eigenvalue = a scalar used in matrix multiplication
* several eigenvalues for a given matrix *i.e. several scalars can work when doing this transformation/matrix multiplication*
* largest, positive eigenvalue of a Leslie matrix *dominant - largest, positive* = $\lambda$ = population growth rate

*really easy to do in R, will do in lab*

### (maybe skip) Example of stage-structured population

* each column = stage *same as Leslie matrix*
* column 1 = rates from seeds to seeds/seedlings/adults
* column 2 = rates from seedlings to seeds/seedlings/adults
* adults to seeds is fecundity

```{r}
stage <- c(0,0.2,0,0,0.1,0.3,50,0,0.8)
matrix(stage,nrow=3,ncol=3)
```